Reparametrization of COM-Poisson Regression Models with Applications in the Analysis of Experimental Data

### 2017-11-14 article.Rnw
Improve case studies section

In this section, we report three illustrative examples of real count
data analysis obtained from experimental research. We considered as
alternative models for the analysis the standard Poisson model, the
COM-Poisson model in two forms (original and mean parametrization) and
Quasi-Poisson model, based on second-moment assumptions. The data sets
and \texttt{R} codes for their analysis are available on on-line
supplements.

@@

This example relates to cotton plants (\textit{Gossypium hirsutum})
submitted to five levels of artificial defoliation (\texttt{des}) and
crossed with five growth stages (\texttt{est}). The purpose of this
research was to study the effect of defoliation levels at different
growth stages of cotton plants on the cotton production, expressed by
the number of bools produced. The study was conducted in a greenhouse
and the experimental design was completely randomized with five
replicates. This data set was analysed by \citet{Zeviani2014} using the
\textit{Gamma-Count} distribution.

Following the results of \citet{Zeviani2014}, we considered the linear
predictor specified by
$$\log(\mu_{ij}) = \beta_0 + \beta_{1j} \textrm{def}_i + \beta_{2j}
\textrm{def}_i^2$$
where $\mu_{ij}$ is the expected number of cotton bolls for the $i$-th
defoliation level ($i=$ 1: 0\%, 2: 25\%, 3: 50\%, 4: 75\% e 5: 100\%)
and $j$-th growth stage ($j$ = 1: vegetative, 2: flower bud, 3: blossom,
4: boll, 5: boll open), that is, we have a second order effect of
defoliation in each growth stage. The parameters estimates and
goodness-of-fit measures for the Poisson, COM-Poisson, COM-Poisson$_\mu$
and quasi-Poisson models are presented in \autoref{tab:coef-cotton}.

@@

The results presented in \autoref{tab:coef-cotton} show that the
goodness-of-fit measures (maximum log-likelihood, AIC and BIC) are quite
similar between COM-Poisson and COM-Poisson$_\mu$. This shows that the
reparametrization does not change the model fit, as expected. The Poisson
model is clearly unsuitable, being overly conservative. The difference
between the log-likelihoods of the Poisson and COM-Poisson$_\mu$ models
is \Sexpr{lrt.ncap[2, 4]}. Comparing this value with the chi-square
distribution, with one degree of freedom, we have these models are
significantly different ($\phi \neq 0$). The estimated value of
precision parameter, $\hat{\phi}$ is \Sexpr{co1[1]}, indicating
underdispersion.

\autoref{tab:coef-cotton} also shows the advantage of COM-Poisson$_\mu$
(reparametrized), since the estimates are quite similar those obtained by
the Poisson model, whereas estimates obtained by the COM-Poisson model
in original parametrization are on a non-interpretable scale. The ratios
between estimates and their respective standard errors for the
COM-Poisson models are very close to ratios obtained by quasi-Poisson
model. However, it is important to note that COM-Poisson model is a
full parametric approach, that is, there is a probability distribution
associate to counts.

In the \autoref{fig:pred-cotton} are presented the observed values and
curves of fitted values with confidence intervals (95\%) as functions of
the defoliation level for each growth stage. The fitted values are same
for both models, but confidence intervals are bigger for Poisson model
because the equidispersion assumption. The results from COM-Poisson
models are consistent with those from the Gamma-Count model, fitted by
\cite{Zeviani2014}, Poisson-Tweedie model, fitted by \cite{Bonat2017}
and alternative COM-Poisson parametrization, fitted by \cite{Huang2017},
in that both methods indicate underdispersion and significant effects of
defoliation for the vegetative, blossom and boll growth stages.

@@

In order to assess the orthogonality between $\bm{\mu}$ and $\phi$ in
the COM-Poisson$_\mu$ parametrization, \autoref{tab:corr-cotton}
presents the empirical correlations between precision and regression
parameters, obtained from the estimated variance and covariance
matrix. The correlations are practically null considering the
COM-Poisson$_\mu$, whereas to the original parametrization they are
quite high. This reflects a better performance of the maximization
algorithm in mean-parametrization model, in this case COM-Poisson$_\mu$
fit was \Sexpr{time12[3] * 100 / time11[3]}\% slower then COM-Poisson
model. It is important to note that the initial values for the BFGS
algorithm are provided by Poisson fitted model, then in the COM-Poisson
mean-parametrized model the initial values are practically the maximum
likelihood estimates and the effort of maximization is on the precision
parameter $\phi$.

@@

The second example is an experiment in a completely randomized block
with treatments in a $5\times 3$ factorial arrangement about soybean
(\emph{Glicine Max}). The aim of this study was to evaluate the effects
of potassium doses (\texttt{K}) applied to soil (0, 0.3, 0.6, 1.2 and
1.8 $\times$ 100mg dm$^{-3}$) and soil moisture (\texttt{umid}) levels
(37.5, 50, 62.5\%) on the soybean production. The experiment was
carried out in a greenhouse, in pots with two plants, and the count
variable measured was the number of grains per pot
\citep{Serafim2012}. \autoref{fig:desc-soy} (left) shows the number of
grains recorded for each combination of potassium dose and moisture
level, it is important to note the indication of a quadratic level of
the potassium levels. Most points in the sample variance vs. sample
means dispersion diagram (right) are above the identity line, suggesting
overdispersion (block effect not yet removed).

@@

For the analysis of this data set we proposed, based on descriptive
analysis \autoref{fig:desc-soy}, the linear predictor specified as a
linear doses for each humidity level and a global quadratic dose effect,
i.e
$$
\log(\mu_{ijk}) = \beta_0 + \gamma_i + \tau_j +
  \beta_{1}\texttt{K}_k + \beta_{2}\texttt{K}_k^2 +
  \beta_{3j}\texttt{K}_k
$$
with $i=$1: block II, 2: block III, 3: block IV e 4: block V; $j=$1:
50\% e 2: 62.5\%; and $k=$1: 0.0, 2: 0.3, 3: 0.6, 4: 1.2, 5: 1.8 100mg
dm$^{-3}$, where $\gamma_i$ is the effect of $i$-th block, $\tau_j$ is
the effect of $j$-th moisture level and $\beta_{3j}$ is the first order
potassium effect (\texttt{K}) for the $j$-th moisture level
(\texttt{umid}). \autoref{tab:coef-soy} presents the estimates, ratio
between estimate and standard error and goodness-of-fit measures for
the alternative models fitted.

The results in \autoref{tab:coef-soy} has similar interpretations to
results presented in \autoref{case-cotton}. The two parametrization of
COM-Poisson presented very similar goodness-of-fit measures and better
than Poisson model. The difference between the log-likelihoods of the
Poisson and COM-Poisson models is \Sexpr{lrt.ngra[2, 4]}, indicating
that $\phi$ is significantly different of zero. With respect to the
regression parameters, the similarities between models are analogous to
the previous section. Both models indicate effects of block, potassium
dose and moisture level, however the Poisson model indicates the effects
with greater significance, because it does not fit the extra
variability.

@@

The infinite sum $Z(\mu, \phi)$ in cases of oversidepersed counts needs
a greater upper bound for the convergence, making the computation of the
likelihood more expansive. In this case, the upper bound was 700. The
number of likelihood computations, on the BFGS algorithm, was
\Sexpr{c21[1]} and \Sexpr{c22[2]} for the original and
mean-parametrization and the time to fit was \Sexpr{time21[3]} and
\Sexpr{time22[3]} seconds respectively. This may be due the
well-behaviour of likelihood function for orthogonal parameters also,
best start values from Poisson model. In \autoref{tab:corr-soy} we
present the empirical correlations between regression and precision
parameter. The correlations are close to zero for the COM-Poisson$_\mu$
parametrization, indicating the empirical orthogonality between $\mu$
and $\phi$ on COM-Poisson$_\mu$ distribution.

@@

The observed counts and fitted curves for each humidity level with
confidence bands are show in \autoref{fig:pred-soy}. The fitted values
are identical for both models, leading to the same conclusions. On the
other hand, confidence bands for the Poisson model are smaller due to
equidispersion assumption. Considering COM-Poisson models and
quasi-Poisson approach, the confidence bands are quite similar, which
shows their flexibility.

@@

The data set comes from an experiment to measure the reproductive
toxicity of a herbicide, nitrofen, on a species of zooplankton
(\textit{Ceriodaphnia dubia}). 50 animals were randomized into batches
of 10 and each batch was put in a solution with a measured concentration
of nitrofen (0, 0.8, 1.6, 2.35 e 3.10 mug$/10^2$litre)
(\texttt{dose}). Then the number of total live offspring was recorded
\cite{Bailer1994}.

@@

For this data set we proposed three linear predictors. We compare the
performance of models using the likelihood ratio tests to select the
systematic part of model. The linear predictors are
\begin{center}
\begin{minipage}{12cm}
Predictor 1: $\log(\mu_i) = \beta_0 + \beta_1 \texttt{dose}_i$\\
Predictor 2: $\log(\mu_i) = \beta_0 + \beta_1 \texttt{dose}_i +
             \beta_2 \texttt{dose}_i^2$\\
Predictor 3: $\log(\mu_i) = \beta_0 + \beta_1 \texttt{dose}_i +
             \beta_2 \texttt{dose}_i^2 + \beta_3 \texttt{dose}_i^3$.
\end{minipage}
\end{center}

@@

\autoref{tab:anova-ovos} summarizes the results of fitted models and
likelihood ratio tests comparing the sequence of predictors. All models
indicate the cubic effect of nitrofen concentration. Considering this
predictor, there is an evidence of equidispersed counts, $\phi$
estimates of COM-Poisson close to zero and $\sigma$ of quasi-Poisson, to
one. It is interesting to note that if we omit the effects of a greater
order on the predictor, the modes show evidence of overdispersion
counts. This exemplifies the discussion about causes of overdispersion
made in \autoref{introduction}. We can also note that the quasi-Poisson
approach, although robust to equisdispersion assumption, shows a higher
descriptive levels ($p$-values) than parametric models, that is, the
tests under parametric models are most powerful than quasi-Poisson
models in the equidispersed case.

In \autoref{tab:coef-ovos} we presented the estimates of the regression
parameters considering the cubic dose effect (third predictor). The
interpretations are similar to the discussed in the others cases
studies, but in this the Poisson model is also suitable for indicating
the significance of the effects. In addition, note that the parameters
of the standard COM-Poisson model are comparable to the others models,
this occurs because it is the particular case $\phi = 0$ which implies
$\lambda = \mu$. \autoref{fig:pred-ovos} shows the number of live
off-springs observed in experiment with fitted curves and confidence
bands for all model strategies adopted.

@@

The fitted values and confidence bands in \autoref{fig:pred-ovos} come
to have a full overlay. This shows that even in case where is not
necessary to estimate the precision parameter $\phi$, but estimating it
does not lead to incorrect analysis.

@@

Finally, in \autoref{tab:corr-ovos} we presented the empirical
correlations between regression and precision parameters. The results
show that even in the particular case ($\phi=0$), the empirical
correlations for the standard models are not zero. For the
reparametrized model, as discussed in the previous sections, the
correlations are practically null.

>>>>

In this section, we report three illustrative examples of count data
analysis. We considered as alternative models for the analysis the
standard Poisson regression model, the COM-Poisson model in the two
forms (original and new parametrization) and the quasi-Poisson
regression model.  The data sets and \texttt{R} codes for their analysis
are available as supplementary material.

@@

This example relates to cotton plants (\textit{Gossypium hirsutum})
submitted to five levels of artificial defoliation (\texttt{des}) and
crossed with five growth stages (\texttt{est}). The main goal of this
study was to assess the effect of defoliation levels at different growth
stages of cotton plants on the cotton production, expressed by the
number of bolls produced. The study was conducted in a greenhouse and
the experimental design was completely randomized with five
replicates. This data set was analyzed by~\citet{Zeviani2014} using the
\textit{Gamma-Count} distribution.

Following~\citet{Zeviani2014}, the linear predictor is given by
$$\log(\mu_{ij}) = \beta_0 + \beta_{1j} \textrm{def}_i + \beta_{2j}
\textrm{def}_i^2$$
where $\mu_{ij}$ is the expected number of cotton bolls for the $i$-th
defoliation level ($i=$ 1: 0\%, 2: 25\%, 3: 50\%, 4: 75\% e 5: 100\%)
and $j$-th growth stage ($j$ = 1: vegetative, 2: flower bud, 3: blossom,
4: boll, 5: boll open), i.e. we have a second order effect of
defoliation in each growth stage. The parameters estimates and
goodness-of-fit measures for the Poisson, COM-Poisson, COM-Poisson$_\mu$
and quasi-Poisson regression models are presented in
\autoref{tab:coef-cotton}.

@@

The results presented in \autoref{tab:coef-cotton} show that the
goodness-of-fit measures (log-likelihood, AIC and BIC) are quite similar
for the COM-Poisson and COM-Poisson$_\mu$ models.  It suggests that the
reparametrization does not change the model fit, as expected.  The
Poisson model is clearly unsuitable, being overly conservative.  The
difference in terms of log-likelihood value from the Poisson to the
COM-Poisson$_\mu$ model was $\Sexpr{lrt.ncap[2, 4]}$, which in turn
suggests the better fit of the COM-Poisson$_{\mu}$ model. A chi-square
test also supports this statement.  Finally, the estimated value of the
dispersion parameter $\hat{\phi} = \Sexpr{co1[1]}$ suggests underdispersion.

Furthermore, results in \autoref{tab:coef-cotton} also shows the
advantage of the COM-Poisson$_\mu$ model, since the estimates are quite
similar to the ones obtained by the Poisson model, whereas estimates
obtained from the COM-Poisson model in the original parametrization are
on a non interpretable scale.  The ratios between estimates and their
respective standard errors for the COM-Poisson models are very close to
ratios obtained by quasi-Poisson model. However, it is important to note
that COM-Poisson model is a full parametric approach, i.e. there is a
probability distribution associate to counts. On the other hand, the
quasi-Poisson model is a specification based only on second-moments
assumptions.

\autoref{fig:pred-cotton} presents the observed and fitted values with
confidence intervals (95\%) as a function of the defoliation level for
each growth stage. The fitted values are the same for the Poisson and
COM-Poisson$_{\mu}$ models, however, the confidence intervals are larger
for the Poisson model because the equidispersion assumption.  The
results from the COM-Poisson$_{\mu}$ model are consistent with those
from the Gamma-Count model~\cite{Zeviani2014},
Poisson-Tweedie~\cite{Bonat2017} and the alternative parametrization of
the COM-Poisson distribution proposed by~\cite{Huang2017}.  In all
strategies the models indicated underdispersion and significant effects
of defoliation for the vegetative, blossom and boll growth stages.

@@

In order to assess the relation between $\bm{\mu}$ and $\phi$ in the
COM-Poisson$_\mu$ parametrization, \autoref{tab:corr-cotton} presents
the empirical correlations between the regression and dispersion
parameters, as computed by the asymptotic covariance matrix of the
estimators, i.e. the inverse of the observed information.  The
correlations are practically null considering the COM-Poisson$_\mu$. On
the other hand, for the original parametrization such correlations are
quite larger, in special for the parameter $\beta_0$ (due to effects
parametrization in the linear predictor).  This result explain the
better performance of the maximization algorithm in the new
parametrization.  It is important to note that the initial values for
the \texttt{BFGS} algorithm are provided by the Poisson model, then in
the COM-Poisson$_{\mu}$ model the initial values are practically the
maximum likelihood estimates and the effort of maximization is on the
dispersion parameter $\phi$ only. To compare the computational times on
the two parametrizations we repeat $50$ times the routine to fit. In
this case COM-Poisson$_\mu$ fit was, on average, $38$\% faster than the
original one.

@@

The second example is an experiment in a completely randomized block
with treatments in a $5\times 3$ factorial arrangement about soybean
(\emph{Glicine Max}). The aim of this study was to evaluate the effects
of potassium doses (\texttt{K}) applied to soil (0, 0.3, 0.6, 1.2 and
1.8 $\times$ 100mg dm$^{-3}$) and soil moisture (\texttt{umid}) levels
(37.5, 50, 62.5\%) on the soybean production. The experiment was carried
out in a greenhouse, in pots with two plants, and the count variable
measured was the number of grains per pot~\citep{Serafim2012}.
\autoref{fig:desc-soy} (left) shows the number of grains recorded for
each combination of potassium dose and moisture level, it is important
to note the indication of a quadratic effect of the potassium
levels. Most points in the sample variance \textit{versus} sample means
dispersion diagram (right) are above the identity line, suggesting
overdispersion (block effect not yet removed).

@@

For the analysis of this data set we proposed based on descriptive
analysis \autoref{fig:desc-soy}, the linear predictor specified as a
linear doses for each humidity level and a global quadratic dose effect,
i.e
$$
\log(\mu_{ijk}) = \beta_0 + \gamma_i + \tau_j +
  \beta_{1}\texttt{K}_k + \beta_{2}\texttt{K}_k^2 +
  \beta_{3j}\texttt{K}_k
$$
with $i=$1: block II, 2: block III, 3: block IV e 4: block V; $j=$1:
50\% e 2: 62.5\%; and $k=$1: 0.0, 2: 0.3, 3: 0.6, 4: 1.2, 5: 1.8 100mg
dm$^{-3}$, where $\gamma_i$ is the effect of $i$-th block, $\tau_j$ is
the effect of $j$-th moisture level and $\beta_{3j}$ is the first order
potassium effect (\texttt{K}) for the $j$-th moisture level
(\texttt{umid}). \autoref{tab:coef-soy} presents the estimates, ratio
between estimate and standard error and goodness-of-fit measures for
the alternative models.

The results in \autoref{tab:coef-soy} have similar interpretations to
the results presented in \autoref{case-cotton}. The two parametrization
of COM-Poisson model presented very similar goodness-of-fit measures and
better fit than the Poisson model. The difference between the
log-likelihoods of the Poisson and COM-Poisson models was
$\Sexpr{lrt.ngra[2, 4]}$, indicating that $\phi$ is significantly
different from zero. Concerning to the regression parameters, the
similarities between the models are analogous to the previous
Section. Both models indicate effects of block, potassium dose and
moisture level, however the Poisson model indicates the effects with
greater significance, because it does not fit the extra variability.

@@

The infinite sum $Z(\mu, \phi)$ in the cases of overdispersed count data
requires a larger upper bound to reach convergence. Thus, in these cases
the computation of the log-likelihood function is computationally
expensive.  For the data set considered, the upper bound was fixed at
$700$.  The \texttt{BFGS} algorithm evaluated the log-likelihood
function $\Sexpr{c21[1]}$ and $\Sexpr{c22[2]}$ times to reach
convergence, when using the original and new parametrization of the
COM-Poisson distribution, respectively.  In terms of computational time
and for 50 repetitions of fit, the proposed reparametrization was on
average $110\%$ faster than the original one.  Probably, it is due to
the better behaviour of the log-likelihood function as well as better
initial values obtained from the Poisson fit. In \autoref{tab:corr-soy}
we present the empirical correlation between the regression and
dispersion parameter estimates. The correlations are close to zero for
the COM-Poisson$_\mu$ model, indicating the empirical orthogonality
between $\mu$ and $\phi$.

@@

The observed and fitted counts for each humidity level with confidence
intervals are shown in \autoref{fig:pred-soy}. The fitted values are
identical for the Poisson and COM-Poisson$_{\mu}$ models, leading to the
same conclusions. On the other hand, confidence intervals for the
Poisson model are smaller than the ones from the
COM-Poisson$_{\mu}$. Probably, it is due to the equidispersion
assumption underline the Poisson model. The confidence intervals from
the COM-Poisson$_{\mu}$ and quasi-Poisson models are really similar,
which in turn show the already highlighted similarity between these
approches.

@@

The data set comes from an experiment to measure the reproductive
toxicity of a herbicide, nitrofen, on a species of zooplankton
(\textit{Ceriodaphnia dubia}). Fifty animals were randomized into
batches of ten and each batch was put in a solution with a measured
concentration of nitrofen (0, 0.8, 1.6, 2.35 e 3.10 mug$/10^2$litre)
(\texttt{dose}). Then the number of total live offspring was recorded
\cite{Bailer1994}.

@@

For this data set we proposed three linear predictors. We compare the
performance of models using the likelihood ratio tests. The linear
predictors are
\begin{center}
\begin{minipage}{12cm}
Predictor 1: $\log(\mu_i) = \beta_0 + \beta_1 \texttt{dose}_i$\\
Predictor 2: $\log(\mu_i) = \beta_0 + \beta_1 \texttt{dose}_i +
             \beta_2 \texttt{dose}_i^2$\\
Predictor 3: $\log(\mu_i) = \beta_0 + \beta_1 \texttt{dose}_i +
             \beta_2 \texttt{dose}_i^2 + \beta_3 \texttt{dose}_i^3$.
\end{minipage}
\end{center}

@@

\autoref{tab:anova-ovos} summarizes the results of the fitted models and
likelihood ratio tests comparing the sequence of predictors. All models
indicate the significance of the cubic effect of the nitrofen
concentration.  Considering this predictor, there is an evidence of
equidispersed counts, $\phi$ estimates of the COM-Poisson close to zero
and $\sigma$ of quasi-Poisson close to one.  It is interesting to note
that if we omit the high order effects the models show evidence of
overdispersion. This exemplifies the discussion about causes of
overdispersion made in \autoref{introduction}.  We can also note that
the quasi-Poisson approach, although robust to equidispersion
assumption, shows a higher descriptive levels ($p$-values) than
parametric models, that is, the tests under parametric models are more
powerful than the ones under the quasi-Poisson model in the
equidispersed case.

In \autoref{tab:coef-ovos} we present the estimates of the regression
parameters considering the cubic dose effect (third predictor). The
interpretations are similar to others cases studies, however, in this
case the Poisson model is also suitable for indicating the significance
of the covariate effects.  In addition, note that the parameter
estimates of the original COM-Poisson model are comparable to the others
models. It occurs because it is the special case $\phi = 0$ which
implies $\lambda = \mu$. \autoref{fig:pred-ovos} shows the number of
live off-springs observed and fitted along with confidence intervals for
all model strategies adopted.

@@

The fitted values and confidence intervals in \autoref{fig:pred-ovos}
come to have a full overlay. It shows that the estimation of the extra
dispersion parameter does not affect the estimation of the regression
coefficients in the case of equidispersed counts.

@@

Finally, in \autoref{tab:corr-ovos} we present the empirical
correlations between the regression and dispersion parameter estimates.
The results show that even in the special case ($\phi=0$), the empirical
correlations for the original COM-Poisson model are not zero.  For the
reparametrized model, as discussed in the previous sections, the
correlations are practically null. The computational times for fifty
repetitions of fit are similar. The average time to fit for the
COM-Poisson$_\mu$ and COM-Poisson models is 1.19 and 1.09 seconds,
respectively.
